machine_learning_analysis

### 2024-11-05 04_shiny_app/Biomarker_Explorer/www/preparing_files.Rmd
Msystems resubmission

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fixing the stats files ##
 protein_to_gene = readr::read_csv("../../../00_r_input_data/protein_id_to_gene.csv")
  gene_description = read_csv("../../../00_r_input_data/all_protein_annotations.csv")
  ## Loading all the data that will be needed ### 
    healthy = read_csv("../../../01_r_processed_data/Infected_vs_Healthy.csv") %>% 
      left_join(protein_to_gene,by = c("uniprotid" = "ProteinID")) %>% 
      left_join(gene_description, by = c("uniprotid" = "Entry")) %>% 
      mutate(url = paste0("https://www.uniprot.org/uniprotkb/",uniprotid,"/entry")) %>% 
      rename(Feature = Gene,
             Description = Protein.names)

@@

## Generating EFS ranking

```{r}
### Healthy vs Infected ###

pl = read_csv("protein_data.csv")

infected_healthy_l = pl %>% 
  mutate(classlabels = case_when(condition %in% c("faecalis","faecium") ~ 1,
                                   TRUE ~ 0))  %>% 
  mutate(classlabels = as.logical(classlabels)) %>% 
  # - messes up EFS for some reason, very annoying. 
  mutate(Feature = gsub("-","_",Feature))

infected_healthy_w = infected_healthy_l %>% 
 tidyr::pivot_wider(names_from = Feature,
              values_from = Abundance) %>% 
  dplyr::select_if(~ !any(is.na(.))) %>% 
  select(-1,-2) %>% 
  #needs to be a dataframe or EFS freaks out. 
  as.data.frame()

set.seed(2)

## Running EFS ##
ensemble_result = EFS::ensemble_fs(infected_healthy_w,classnumber = 1,cor_threshold = 0)

## Getting EFS Result in Plotable Format
long = ensemble_result %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("model") %>% 
  pivot_longer(2:length(.),names_to = "Gene") %>% 
  #converting genes back to proper notation. 
  mutate(Gene = gsub("_","-",Gene)) %>% 
  rename(Feature = Gene)

# Ranking Biomarkers
top = long %>% 
  group_by(Feature) %>% 
  dplyr::summarise(sum = sum(value)) %>% 
  dplyr::arrange(-sum) %>% 
  tibble::rowid_to_column(var = "rank")

write_csv(top,"healthy_prot_EFS_rank.csv")

### Faecalis vs Faecium ###

fvf = pl %>% 
  filter(condition != "healthy") %>% 
  mutate(classlabels = case_when(condition %in% c("faecalis") ~ 1,
                                   TRUE ~ 0)) %>% 
  mutate(classlabels = as.logical(classlabels)) %>% 
  # - messes up EFS for some reason, very annoying. 
  mutate(Feature = gsub("-","_",Feature))

fvf_w = fvf %>% 
 tidyr::pivot_wider(names_from = Feature,
              values_from = Abundance) %>% 
  dplyr::select_if(~ !any(is.na(.))) %>% 
  select(-1,-2) %>% 
  #needs to be a dataframe or EFS freaks out. 
  as.data.frame()

set.seed(2)

## Running EFS ##
ensemble_result = EFS::ensemble_fs(fvf_w,classnumber = 1,cor_threshold = 0)

## Getting EFS Result in Plotable Format
long = ensemble_result %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("model") %>% 
  pivot_longer(2:length(.),names_to = "Gene") %>% 
  #converting genes back to proper notation. 
  mutate(Gene = gsub("_","-",Gene)) %>% 
  rename(Feature = Gene)

# Ranking Biomarkers
top = long %>% 
  group_by(Feature) %>% 
  dplyr::summarise(sum = sum(value)) %>% 
  dplyr::arrange(-sum) %>% 
  tibble::rowid_to_column(var = "rank")

write_csv(top,"EcB_prot_EFS_rank.csv")

### Mortality vs Survival ### 
mvs = pl %>% 
  filter(!is.na(death_during_admission)) %>% 
  # - messes up EFS for some reason, very annoying. 
  mutate(Feature = gsub("-","_",Feature))

mvs_w = mvs %>% 
 tidyr::pivot_wider(names_from = Feature,
              values_from = Abundance) %>% 
  dplyr::select_if(~ !any(is.na(.))) %>% 
  select(-1,-2) %>% 
  #needs to be a dataframe or EFS freaks out. 
  as.data.frame()

set.seed(2)

## Running EFS ##
ensemble_result = EFS::ensemble_fs(mvs_w,classnumber = 1,cor_threshold = 0)

## Getting EFS Result in Plotable Format
long = ensemble_result %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("model") %>% 
  pivot_longer(2:length(.),names_to = "Gene") %>% 
  #converting genes back to proper notation. 
  mutate(Gene = gsub("_","-",Gene)) %>% 
  rename(Feature = Gene)

# Ranking Biomarkers
top = long %>% 
  group_by(Feature) %>% 
  dplyr::summarise(sum = sum(value)) %>% 
  dplyr::arrange(-sum) %>% 
  tibble::rowid_to_column(var = "rank")

write_csv(top,"mortality_prot_EFS_rank.csv")

```
```{r}

      vp %>% ggboxplot(x = "condition",
                      y = "Abundance",
                      facet.by = "Feature",
                      scales = "free")+
        #facet_wrap(~condition, scales = "free") +
        stat_pvalue_manual(stat,label = "p.signif")+
        ggprism::theme_prism()
```

@@

```{r}
gnps_annotations = read_csv("../../../../Manuscript_v2/00_r_input_data/gnps_annotations.csv")
normalized_data = read_csv("normalized_metabolomics.csv")

@@

## Generating EFS Ranking for Metabolomics

normalized_data = read_csv("../../../01_r_processed_data/normalized_metabolomics.csv")

### Infected vs Healthy ### 
## Switching to the wide format ### 
norm_wide = normalized_data %>%
  select(-row_m_z,-row_retention_time,-value) %>% 
  tidyr::pivot_wider(names_from = row_id,values_from = norm_value)

## Looking for only columns that were annotated ## 
gnps_and_in_norm = intersect(unique(gnps_annotations$row_id), colnames(norm_wide))

gnps_annotated_metabolites = norm_wide %>% 
  as.data.frame() %>% 
  select(condition,gnps_and_in_norm)

## Extracting the numeric data ##
efs_sel = gnps_annotated_metabolites %>% 
  mutate(classlabels = case_when(condition %in% c("faecalis","faecium") ~ 0,
                           TRUE ~ 1)) %>% 
  select(classlabels,2:length(.)) %>% 
   #absolutely needs to be a data frame and not a tibble for EFS package
  as.data.frame()

# Seting seed incase this isn't done internally in EFS
set.seed(2)

## Running EFS feature selection ##
ensemble_result = EFS::ensemble_fs(efs_sel,
                                   classnumber = 1,cor_threshold = 0)

## Getting EFS Result in Plotable Format
long = ensemble_result %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("model") %>% 
  pivot_longer(2:length(.),names_to = "row_id") %>% 
  left_join(gnps_annotations,by = "row_id") %>% 
  rename(Feature = row_id)

# Ranking Biomarkers
top = long %>% 
  group_by(Feature,compound_name_cleaned) %>% 
  dplyr::summarise(sum = sum(value)) %>% 
  dplyr::arrange(-sum) %>% 
  tibble::rowid_to_column(var = "rank") %>% 
  select(rank,Feature,compound_name_cleaned,sum)

write_csv(top,"infected_metabolomics_EFS.csv")
```

Faecalis vs Faecium 
```{r}

normalized_data = read_csv("../../../01_r_processed_data/normalized_metabolomics.csv")

## Reading in the normalized data ## 
normalized_data = normalized_data %>% 
#Removing reduntant RIDs. These were found to be the same as other top biomarkers
  filter(row_id != "RID_8856",
         row_id != "RID_22529",
         row_id != "RID_22270",
         row_id != "RID_9436")

## Switching to the wide format ### 
norm_wide = normalized_data %>%
  select(-row_m_z,-row_retention_time,-value) %>% 
  tidyr::pivot_wider(names_from = row_id,values_from = norm_value)

## Looking for only columns that were annotated ## 
gnps_and_in_norm = intersect(unique(gnps_annotations$row_id), colnames(norm_wide))

gnps_annotated_metabolites = norm_wide %>% 
  as.data.frame() %>% 
  select(condition,gnps_and_in_norm)

## Extracting the numeric data ##
efs_sel = gnps_annotated_metabolites %>% 
  filter(condition %in% c("faecalis","faecium")) %>% 
  mutate(classlabels = case_when(condition == "faecalis" ~ 0,
                           TRUE ~ 1)) %>% 
  select(classlabels,2:length(.)) %>% 
   #absolutely needs to be a data frame and not a tibble for EFS package
  as.data.frame()

# Seting seed incase this isn't done internally in EFS
set.seed(2)

## Running EFS feature selection ##
ensemble_result = EFS::ensemble_fs(efs_sel,
                                   classnumber = 1,
                                   cor_threshold = 0)

## Getting EFS Result in Plotable Format
long = ensemble_result %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("model") %>% 
  pivot_longer(2:length(.),names_to = "row_id") %>% 
  left_join(gnps_annotations,by = "row_id") %>% 
  rename(Feature = row_id)

# Ranking Biomarkers
top = long %>% 
  group_by(Feature,compound_name_cleaned) %>% 
  dplyr::summarise(sum = sum(value)) %>% 
  dplyr::arrange(-sum) %>% 
  tibble::rowid_to_column(var = "rank") %>% 
  select(rank,Feature,compound_name_cleaned,sum)

write_csv(top,"EcB_metabolomics_EFS.csv")

Mortality

```{r}
normalized_data = read_csv("../../../01_r_processed_data/normalized_metabolomics.csv")

# Switching to the wide format ### 
norm_wide = normalized_data %>%
  select(-row_m_z,-row_retention_time,-value) %>% 
  tidyr::pivot_wider(names_from = row_id,values_from = norm_value)

## Looking for only columns that were annotated ## 
gnps_and_in_norm = intersect(unique(gnps_annotations$row_id), colnames(norm_wide))

gnps_annotated_metabolites = norm_wide %>% 
  as.data.frame() %>% 
  select(death_during_admission,gnps_and_in_norm) %>% 
  filter(!is.na(death_during_admission))

# Seting seed incase this isn't done internally in EFS
set.seed(2)

## Running EFS feature selection ##
ensemble_result = EFS::ensemble_fs(gnps_annotated_metabolites,
                                   classnumber = 1,cor_threshold = 0)

## Getting EFS Result in Plotable Format
long = ensemble_result %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("model") %>% 
  pivot_longer(2:length(.),names_to = "row_id") %>% 
  left_join(gnps_annotations,by = "row_id") %>% 
  rename(Feature = row_id)

# Ranking Biomarkers
top = long %>% 
  group_by(Feature,compound_name_cleaned) %>% 
  dplyr::summarise(sum = sum(value)) %>% 
  dplyr::arrange(-sum) %>% 
  tibble::rowid_to_column(var = "rank") %>% 
  select(rank,Feature,compound_name_cleaned,sum)

write_csv(top,"mortality_metabolomics_EFS.csv")

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(ggpubr)
library(tidyverse)
```

```{r}
## Fixing the stats files ##
 protein_to_gene = readr::read_csv("../../../00_r_input_data/protein_id_to_gene.csv")
  gene_description = read_csv("../../../00_r_input_data/all_protein_annotations.csv")
  ## Loading all the data that will be needed ### 
    healthy = read_csv("../../../01_r_processed_data/Infected_vs_Healthy.csv") %>% 
      left_join(protein_to_gene,by = c("uniprotid" = "ProteinID")) %>% 
      left_join(gene_description, by = c("uniprotid" = "Entry")) %>% 
      mutate(url = paste0("https://www.uniprot.org/uniprotkb/",uniprotid,"/entry")) %>% 
      rename(Feature = Gene,
             Description = Protein.names)

@@

@@

```{r}
gnps_annotations = read_csv("../../../00_r_input_data/gnps_annotations.csv")
normalized_data = read_csv("../../../01_r_processed_data/normalized_metabolomics.csv")

@@

### 2024-11-05 08_Msystems_resubmission/00_machine_learning_analysis/machine_learning_analysis.qmd
Msystems resubmission

1. Healthy vs Infected patients
2. E.faecalis vs E.faecium infected patients
3. Mortality vs Surival.

@@

We will split our

@@

@@

# Defining the model Max built as a function, so we can easily apply it to all 
# of our comparisons.
our_model = function(model_data,
                     outcome,
                     model_recipe){
# Setting seed at 42 to make reproducible.  
set.seed(42)
# Create train/test split
train_test_split <- initial_split(model_data,prop = 0.8,
                                  strata = outcome)
train_data <- training(train_test_split)
test_data <- testing(train_test_split)

@@

  # Plotting coefficients (Lasso-selected features)
  p1 <- model_summary_out %>% 
    ggplot(aes(reorder(term, estimate), estimate)) +
    geom_point() +
    coord_flip()

@@

h_v_i_res$model_summary_plot
```
**Figure 2.** Here we see the features that are important for the model. Several of these features are likely artifacts of the way the samples were collected. Phtalic anhydride and phthalate are plasticizers. These are almost certainly due to the fact that these samples were collected in different types fo plastic tubes. This was a point the reviewers brought up.

@@

```{r}
# Recipe does all of the feature engineering so that we don't have to do anything else to the input data
h_v_i_model_recipe2 <- recipe(infection_status ~ ., data = model_data) %>%
  update_role(sample_id, new_role = "id") %>% 
  update_role(condition, new_role = "id") %>% 
  update_role(`RID_29112:bis(2-ethylhexyl) phthalate` , new_role = "id") %>% 
  update_role(`RID_22644:phthalic anhydride` , new_role = "id") %>% 
  update_role(death_during_admission, new_role = "id") %>%  # id role tells the model not to actually incorporate it as a feature
  update_role(infection_status, new_role = "outcome") %>% # specify target 
  step_dummy(all_nominal_predictors()) %>% # convert string to dummy variables
  step_normalize(all_numeric_predictors()) # normalize numeric features
```

```{r}
h_v_i_res2 <- our_model(model_data,"infection_status",h_v_i_model_recipe2)

extract_roc(h_v_i_res2,"infection_status",".pred_infected")
```
**Figure 3.** Model still performs perfectly with problematic metabolites removed.

```{r}
h_v_i_res2$model_summary_plot
```

@@

```{r}
#| label: model ROCs
extract_roc(faecalis_v_faecium_res,"condition",".pred_faecalis")
```

@@

```{r}
#| label: model estimates
#| 
faecalis_v_faecium_res$model_summary_plot
```
**Figure 6.** Here we can see the features that are most important to the model. Looks very similar to the previous analysis. Several of these proteins and metabolites are immune related.

@@

```{r}
model_data <- clinical_metadata %>% 
  filter(condition != "healthy") %>% 
  # antimicrobial testing was not perfomred for this patient
  filter(sample_id != "S18") %>% 
  #getting rid of uninformative columns. Either very little variance, or too many levels to be useful 
  select(-pathogen,-patient_id,-pathogen_source,-infection_status,
         -bacteremia_recurrence,-race,-death_at_one_year,
         -peripheral_vascular_disease,-dementia,-hemiplegia,
         -highly_active_antiretroviral_therapy,
         -source) %>% 
  mutate(transplant_type = case_when(transplant_type == "none" ~ "none",
                                     TRUE ~ "transplant")) %>% 
  select_if(~ !any(is.na(.))) %>% 
  select(-antibiotic_therapy_before_48_hrs,-antibiotic_therapy_after_48_hrs) %>% 
   mutate(across(c(history_of_myocardial_infarction, 
                   congestive_heart_failure, 
                   history_of_stroke_or_transient_ischemic_attack, 
                   smoking, copd, peptic_ulcer_disaese, 
                   mod_to_severe_chromic_kidney_disease, 
                   hodgkin_disease, leukemia_or_lymphoma, 
                   icu_admission, mechanical_ventilation, 
                    polymicrobial_bacteremia, 
                   thrombocytopenia_plt_50, hypotension,sensitive_to_vancomycin), 
                as.factor)) %>% 
  mutate(wbc = as.numeric(wbc))

@@

extract_roc(condition_metadata_res,"condition",".pred_faecalis")
```
**Figure 7.** The model based on the metadata alone performs well. It actually performs better than our metabolomics/proteomics model.

@@

```{r}
condition_metadata_res$model_summary_plot
```
**Figure 8.** Here we can see that the model values amp_mic as the most important features.

@@

extract_roc(condition_metadata_res2,"condition",".pred_faecalis")
```
**Figure 8.** When we do this, we find that that model performs slightly worse than our metabolomics/proteomics model, but on the same scale.

```{r}
condition_metadata_res2$model_summary_plot
```
**Figure 9.** When we look into what is driving this model, we see that whether or not the patients had a transplant is a strong predictor.

This is not novel, and highlights how if you are immunocompromised, you are more likely to get a E.faecium infection.

What if we take out transplant type from the model?
```{r}

faecalis_v_faecium_m_model_recipe3 <- recipe(condition ~ ., data = model_data) %>%
  update_role(sample_id, new_role = "id") %>% 
  update_role(death_during_admission, new_role = "id") %>% 
  update_role(pathogen_type,new_role = "id") %>% 
  update_role(sensitive_to_vancomycin,new_role = "id") %>% 
  update_role(amp_mic,new_role = "id") %>% 
  update_role(vanc_mic,new_role = "id") %>% 
  update_role(transplant_type, new_role = "id")%>%  # id role tells the model not to actually incorporate it as a feature
  update_role(condition, new_role = "outcome") %>% # specify target 
  step_novel() %>% 
  step_dummy(all_nominal_predictors()) %>% # convert string to dummy variables
  step_impute_knn(wbc) %>% 
  step_normalize(all_numeric_predictors()) # normalize numeric features

condition_metadata_res3 = our_model(model_data,"condition",
                                   faecalis_v_faecium_m_model_recipe3)

extract_roc(condition_metadata_res3,"condition",".pred_faecalis")
```
**Figure 10.** Here we can see once all of these variables are taken out, the metadata model is essentially worthless.

All this to say, we did not find any clinical variables that are novel and 
predictive.

```{r}
condition_metadata_res3$model_summary_plot
```
**Figure 11.** Here are the features that are important for the worthless model.

@@

```{r}
extract_roc(mortality_vs_survival_res,"death_during_admission",".pred_FALSE")
```
**Figure 12.** Here is a ROC curve of our model. It looks like it is reasonably 
accurate, but note that there is a big class imbalance here. A simple model 
where no patients die would get it right 76% of the time (57/75). 
Our fancy model is only slightly slightly improved on this. 
We are really hindered by our ability to only test a total of 3 true positives 
(due to the small sample size).

```{r}
mortality_vs_survival_res$confusion_matrix
```

@@

```{r}
mortality_vs_survival_res$model_summary_plot
```
**Figure 13.** Here we can see the features important for the model. Note the size of the intercept.

@@

```{r}
model_data <- clinical_metadata %>% 
  filter(condition != "healthy") %>% 
  # antimicrobial testing was not perfomred for this patient
  filter(sample_id != "S18") %>% 
  #getting rid of uninformative columns. Either very little variance, or too many levels to be useful 
  select(-pathogen,-patient_id,-pathogen_source,-infection_status,
         -bacteremia_recurrence,-race,-death_at_one_year,
         -peripheral_vascular_disease,-dementia,-hemiplegia,
         -highly_active_antiretroviral_therapy,
         -source) %>% 
  mutate(transplant_type = case_when(transplant_type == "none" ~ "none",
                                     TRUE ~ "transplant")) %>% 
  select_if(~ !any(is.na(.))) %>% 
  select(-antibiotic_therapy_before_48_hrs,-antibiotic_therapy_after_48_hrs) %>% 
   mutate(across(c(history_of_myocardial_infarction, 
                   congestive_heart_failure, 
                   history_of_stroke_or_transient_ischemic_attack, 
                   smoking, copd, peptic_ulcer_disaese, 
                   mod_to_severe_chromic_kidney_disease, 
                   hodgkin_disease, leukemia_or_lymphoma, 
                   icu_admission, mechanical_ventilation, 
                    polymicrobial_bacteremia, 
                   thrombocytopenia_plt_50, hypotension,sensitive_to_vancomycin,
                   death_during_admission), 
                as.factor)) %>% 
  mutate(wbc = as.numeric(wbc))

@@

extract_roc(mort_metadata_res,"death_during_admission",".pred_FALSE")
```
**Figure 14.** Here we see that the clinical metadata does better job of predicting mortality than oru

@@

```{r}
mort_metadata_res$model_summary_plot
```
**Figure 15.* Turns out if you get admitted to ICU, that predicts mortality 
pretty well (obviously). Once again, this is kind of cheating, ICU admission is 
a feature that is only apparent after a patient is at the hospital for some time.

@@

```{r}
mort_m_model_recipe <- recipe(death_during_admission ~ ., data = model_data) %>%
  update_role(sample_id, new_role = "id") %>% 
  update_role(death_during_admission, new_role = "outcome") %>%  # id role tells the model not to actually incorporate it as a feature
  update_role(icu_admission, new_role = "id") %>% # specify target 
  step_dummy(all_nominal_predictors()) %>% # convert string to dummy variables
  step_impute_knn(wbc) %>% 
  step_normalize(all_numeric_predictors()) # normalize numeric features

@@

extract_roc(mort_metadata_res,"death_during_admission",".pred_FALSE")

```
**Figure 16.** If you take out the ICU admission, you lose all model performance.
You actually would be better off just predicting no one dies.

>>>>

1.  Healthy vs Infected patients
2.  E.faecalis vs E.faecium infected patients
3.  Mortality vs Survival.

@@

@@

```{r}
#| echo: false
#| eval: false
# Cleaning the data and saving to file for easy reference.

proteomics = read_csv("../../01_r_processed_data/quant_msstats_protein_level_data.csv") %>% 
  dplyr::select(-(1:4)) %>% 
  dplyr::select(-sample_id,-Condition) %>% 
  tidyr::pivot_wider(names_from = BioReplicate,values_from = Abundance) %>% 
  dplyr::filter(orientation == "sp") %>% 
  na.omit()

write_csv(proteomics,"proteomics_data_cleaned.csv")

metabolo_id = read_csv("../../00_r_input_data/gnps_annotations.csv") %>% 
  select(row_id,compound_name = compound_name_cleaned)

metabolomics = read_csv("../../01_r_processed_data/normalized_metabolomics.csv") %>% 
  select(row_id,sample_id,norm_value) %>% 
  left_join(metabolo_id,by = "row_id") %>% 
  pivot_wider(names_from = sample_id,values_from = norm_value)

write_csv(metabolomics, "metabolomics_data_cleaned.csv")

@@

# Defining the model Max built as a function, so we can easily apply it to all 
# of our comparisons.
our_model = function(model_data,
                     outcome,
                     model_recipe){
# Setting seed at 2 to make reproducible.  
set.seed(2)
# Create train/test split
train_test_split <- initial_split(model_data,prop = 0.8,
                                  strata = outcome)
train_data <- training(train_test_split)
test_data <- testing(train_test_split)

@@

  # Plotting coefficients (Lasso-selected features)
  p1 <- model_summary_out %>% 
    ggplot(aes(reorder(term, estimate), estimate)) +
    geom_point() +
    coord_flip()+
    xlab("term")

@@

h_v_i_res$model_summary_plot
```

**Figure 2.** Here we see the features that are important for the model. Several of these features are likely artifacts of the way the samples were collected. Phtalic anhydride and phthalate are plasticizers. These are almost certainly due to the fact that these samples were collected in different types fo plastic tubes. This was a point the reviewers brought up.

@@

```{r}
# Recipe does all of the feature engineering so that we don't have to do anything else to the input data
h_v_i_model_recipe2 <- recipe(infection_status ~ ., data = model_data) %>%
  update_role(sample_id, new_role = "id") %>% 
  update_role(condition, new_role = "id") %>% 
  update_role(`RID_29112:bis(2-ethylhexyl) phthalate` , new_role = "id") %>% 
  update_role(`RID_21438:decynediol, tetramethyl, di(2-hydroxyethyl) ether` , new_role = "id") %>%            update_role(`RID_10819:pentapropylene glycol` , new_role = "id") %>% 
  update_role(`RID_22644:phthalic anhydride` , new_role = "id") %>% 
  update_role(death_during_admission, new_role = "id") %>%  # id role tells the model not to actually incorporate it as a feature
  update_role(infection_status, new_role = "outcome") %>% # specify target 
  step_dummy(all_nominal_predictors()) %>% # convert string to dummy variables
  step_normalize(all_numeric_predictors()) # normalize numeric features
```

```{r,fig.height =3, fig.width= 3}
h_v_i_res2 <- our_model(model_data,"infection_status",h_v_i_model_recipe2)

p1= extract_roc(h_v_i_res2,"infection_status",".pred_infected")

p1

ggsave("healthy_vs_infected_ROC.pdf",p1,width = 3,height = 3)

ggsave("../../03_supplemental/s11_machine_learning_proteomics_metabolomics/S11_a_1.pdf",p1,width = 3,height = 3)
```

**Figure 3.** Model still performs perfectly with problematic metabolites removed.

```{r,fig.height = 3.5, fig.width = 7}
p2 = h_v_i_res2$model_summary_plot

p2

ggsave("healthy_vs_infected_estimates.pdf",p2, height = 3.5, width = 7 )
ggsave("../../03_supplemental/s11_machine_learning_proteomics_metabolomics/S11_a_2.pdf",p2,width = 7,height = 3.5)
```

@@

```{r,fig.height = 3, fig.width = 3}
#| label: model ROCs
p3 = extract_roc(faecalis_v_faecium_res,"condition",".pred_faecalis")

p3

ggsave("faecalis_v_faecium_ROC.pdf",p3, height= 3, width = 3)

ggsave("../../03_supplemental/s11_machine_learning_proteomics_metabolomics/S11_b_1.pdf",p3,width = 3,height = 3)
```

@@

```{r,fig.height = 3.5, fig.width = 7 }
#| label: model estimates
#| 
p4 = faecalis_v_faecium_res$model_summary_plot
p4

ggsave("faecalis_vs_faecium_estimates.pdf",p4, height = 3.5, width = 7 )
ggsave("../../03_supplemental/s11_machine_learning_proteomics_metabolomics/S11_b_2.pdf",p4,width = 7,height = 3.5)
```

**Figure 6.** Here we can see the features that are most important to the model. Looks very similar to the previous analysis. Several of these proteins and metabolites are immune related.

@@

```{r,fig.height = 3, fig.width = 3}
model_data <- clinical_metadata %>% 
  filter(condition != "healthy") %>% 
  # antimicrobial testing was not perfomred for this patient
  filter(sample_id != "S18") %>% 
  #getting rid of uninformative columns. Either very little variance, or too many levels to be useful 
  select(-pathogen,-patient_id,-pathogen_source,-infection_status,
         -bacteremia_recurrence,-race,-death_at_one_year,
         -peripheral_vascular_disease,-dementia,-hemiplegia,
         -highly_active_antiretroviral_therapy,
         -source) %>% 
  mutate(transplant_type = case_when(transplant_type == "none" ~ "none",
                                     TRUE ~ "transplant")) %>% 
  select_if(~ !any(is.na(.))) %>% 
  select(-antibiotic_therapy_before_48_hrs,-antibiotic_therapy_after_48_hrs) %>% 
   mutate(across(c(history_of_myocardial_infarction, 
                   congestive_heart_failure, 
                   history_of_stroke_or_transient_ischemic_attack, 
                   smoking, copd, peptic_ulcer_disaese, 
                   mod_to_severe_chromic_kidney_disease, 
                   hodgkin_disease, leukemia_or_lymphoma, 
                   icu_admission, mechanical_ventilation, 
                    polymicrobial_bacteremia, 
                   thrombocytopenia_plt_50, hypotension,sensitive_to_vancomycin), 
                as.factor)) %>% 
  mutate(wbc = as.numeric(wbc))

@@

p5 = extract_roc(condition_metadata_res,"condition",".pred_faecalis")

p5

ggsave("faecalis_vs_faecium_md_roc_1.pdf",p5, height = 3, width = 3)
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12_a_1.pdf",p5, width = 3,height = 3)
```

**Figure 7.** The model based on the metadata alone performs well. It actually performs better than our metabolomics/proteomics model.

@@

```{r,fig.height = 3.5, fig.width = 5.5}
p6 = condition_metadata_res$model_summary_plot

p6

ggsave("faecalis_vs_faecium_md_estimate_1.pdf",p6, height = 3.5, width = 5.5)
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12_a_2.pdf",p6,
       width = 5.5,
       height = 3.5)
```

**Figure 8.** Here we can see that the model values amp_mic as the most important features.

@@

p7 = extract_roc(condition_metadata_res2,"condition",".pred_faecalis")

p7
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12_b_1.pdf",p7, width = 3 ,height = 3)

ggsave("faecalis_vs_faecium_md_roc_2.pdf",p7, height = 3, width = 3)

p8 = condition_metadata_res2$model_summary_plot
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12_b_2.pdf",p8, width = 5 ,height = 3.5)

**Figure 8.** When we do this, we find that that model performs similarily to our proteomics/metabolomics model

@@

```{r,fig.height = 3, fig.width = 3}
p11 = extract_roc(mortality_vs_survival_res,"death_during_admission",".pred_FALSE")

p11

ggsave("mortality_vs_survival_ROC.pdf",p11, height = 3, width = 3)
ggsave("../../03_supplemental/s11_machine_learning_proteomics_metabolomics/S11_c_1.pdf",p11, width = 3 ,height = 3)

**Figure 12.** Here is a ROC curve of our model. It looks like it is reasonably accurate, but note that there is a big class imbalance here. A simple model where no patients die would get it right 76% of the time (57/75). Our fancy model is only slightly slightly improved on this. We are really hindered by our ability to only test a total of 3 true positives (due to the small sample size).

```{r}
p12 = mortality_vs_survival_res$confusion_matrix

p12

@@

```{r,fig.height = 3.5, fig.width = 8 }
p13 = mortality_vs_survival_res$model_summary_plot

p13

ggsave("mortality_vs_survival_estimate.pdf",p13, height = 3.5, width = 6)
ggsave("../../03_supplemental/s11_machine_learning_proteomics_metabolomics/S11_c_2.pdf",p13, width = 8 ,height = 3.5)
```

**Figure 13.** Here we can see the features important for the model. Note the size of the intercept.

@@

```{r,fig.height = 3, fig.width = 3}
model_data <- clinical_metadata %>% 
  filter(condition != "healthy") %>% 
  # antimicrobial testing was not perfomred for this patient
  filter(sample_id != "S18") %>% 
  #getting rid of uninformative columns. Either very little variance, or too many levels to be useful 
  select(-pathogen,-patient_id,-pathogen_source,-infection_status,
         -bacteremia_recurrence,-race,-death_at_one_year,
         -peripheral_vascular_disease,-dementia,-hemiplegia,
         -highly_active_antiretroviral_therapy,
         -source) %>% 
  mutate(transplant_type = case_when(transplant_type == "none" ~ "none",
                                     TRUE ~ "transplant")) %>% 
  select_if(~ !any(is.na(.))) %>% 
  select(-antibiotic_therapy_before_48_hrs,-antibiotic_therapy_after_48_hrs) %>% 
   mutate(across(c(history_of_myocardial_infarction, 
                   congestive_heart_failure, 
                   history_of_stroke_or_transient_ischemic_attack, 
                   smoking, copd, peptic_ulcer_disaese, 
                   mod_to_severe_chromic_kidney_disease, 
                   hodgkin_disease, leukemia_or_lymphoma, 
                   icu_admission, mechanical_ventilation, 
                    polymicrobial_bacteremia, 
                   thrombocytopenia_plt_50, hypotension,sensitive_to_vancomycin,
                   death_during_admission), 
                as.factor)) %>% 
  mutate(wbc = as.numeric(wbc))

@@

p14 = extract_roc(mort_metadata_res,"death_during_admission",".pred_FALSE")

p14

ggsave("mortality_vs_survival_md_roc_1.pdf",p14,height = 3, width = 3)
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12c_1.pdf",p14,height = 3, width = 3)
```

**Figure 14.** Here we see that the clinical metadata does better job of predicting mortality than oru

@@

```{r,fig.height = 3.5, fig.width= 5.5}
p15 = mort_metadata_res$model_summary_plot

p15

ggsave("mortality_vs_survival_md_estimate_1.pdf",p15, height = 3.5, width = 5.5)
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12c_2.pdf",p15,height = 3, width = 5.5)
```

\*\*Figure 15.\* Turns out if you get admitted to ICU, that predicts mortality pretty well (obviously). Once again, this is kind of cheating, ICU admission is a feature that is only apparent after a patient is at the hospital for some time.

@@

```{r,fig.height = 3, fig.width = 3 }
mort_m_model_recipe <- recipe(death_during_admission ~ ., data = model_data) %>%
  update_role(sample_id, new_role = "id") %>% 
  update_role(death_during_admission, new_role = "outcome") %>%  # id role tells the model not to actually incorporate it as a feature
  update_role(icu_admission, new_role = "id") %>% # specify target 
  step_dummy(all_nominal_predictors()) %>% # convert string to dummy variables
  step_impute_knn(wbc) %>% 
  step_normalize(all_numeric_predictors()) # normalize numeric features

@@

p16 = extract_roc(mort_metadata_res,"death_during_admission",".pred_FALSE")

p16

ggsave("mortality_vs_survival_md_roc_2.pdf",p16, height = 3, width = 3)
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12d_1.pdf",p16,height = 3, width = 3)

**Figure 16.** If you take out the ICU admission, you lose all model performance. You actually would be better off just predicting no one dies.

```{r}
p15 = mort_metadata_res$model_summary_plot

p15

ggsave("mortality_vs_survival_md_estimate_2.pdf",p15, height = 3.5, width = 5.5)
ggsave("../../03_supplemental/s12_machine_learning_clinical_metadata/S12d_2.pdf",p15,height = 3, width = 5.5)
```
